Some lessons after the formalization of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-05-27 Independence_CH/lessons.tex
Removed an extra \end{enumerate}
More lessons into slightly-sensical paragraphs

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example, we only proved the Delta System Lemma
for $\aleph_1$-sized families (thus limiting us to the plain ccc) and
showed preservation of sequences by considering countably closed
forcings (in fact, we formalized the bare minimum requirement of being
$(<\omega+1)$-closed). In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available.

@@

\begin{enumerate}
\item Bureaucracy vs ML programming.
\item The “math” was already formalized on 22 November 2020.
  We finished the last goal on 22 August 2021.
  (Update: 20 November 2021 \& 28 November 2021, for CH)
\item Missing: automation of closure of models under operations.
\item Missing: basic arithmetic for dealing with arities.
\end{enumerate}

\begin{enumerate}
\item It is extremely misleading when automatic tools (\isatt{simp}, \isatt{auto}, etc)
  stop working just because of the sheer size of the goal. Oftentimes,
  in math, we disregard scale issues but they must always be taken
  into account in CS.
\item Example: $\forceisa(0\in 1)$ is expandable,
  $\forceisa(\neg\neg  0\in 1)$ is not.
\item Example: Synthesis of $\forceisa$; could have been fully synthesized,
  but that was dirty “strategy”.
\item The know-how of computer scientists on this kind of engineering is
  very important
\end{enumerate}

\subsection{You might have formalized it, and still be wrong}
\begin{enumerate}
\item Example: restriction of relations.
\item Pollack, “Pollack consistency” by
  Wiedijk. Cf. \theory{Definitions\_Main} (thanks to discussions with
  Vidnyánszky). Opacity of automated proofs. 
\item Plot twist: You can be right without knowing. Intuition may drive proofs
  even if we are not working on what we believe we are.
\end{enumerate}

\subsection{Beware of the “Code fever”}\label{sec:beware-code-fever}
\begin{enumerate}
\item “We know that doing math is fun---formalization is like DRUGS”
\item Feeling of accomplishment after seeing your writings
  validated beyond reasonable doubt (v.g. cofinality).

\item One easily forgets about the “Power of the Board.”
\end{enumerate}

\subsection{The Devil's on the shortcuts}
\begin{enumerate}
\item
  Our proofs of the “definition of forces” (and many
  consequences) and of the lemma for “forcing a value” of function
  depend on the countability of the ground model. 
\item
  Density arguments (look for “TODO”, “general versions”).
\end{enumerate}

\end{enumerate}

>>>>

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families
  (thus limiting us to the plain ccc);
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $(<\omega+1)$-closed); and
\item we proved that appropriate subsets were dense only for Cohen posets
  relevant for the arguments at hand, but could have been done in general.
\end{itemize}
In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isatt{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

\subsection{How to believe in the formalization}
This is a rather tricky question, that was addressed by Pollack in his
\cite{MR1686867}. There is little point to discuss that, after an
assistant has accepted some input successfully, \emph{some
mathematics} has been formally verified. What might not be apparent is
if the claimed theorems are indeed the results that have been
checked. One key aspect of this is the logical foundation of the
assistant (Section~\ref{sec:isabelle-metalogic-meta}). But the weakest
link in the chain is the laying down of definitions building up to the
concepts needed to state the target results. (Another related aspect,
concerning the way results are printed and parsed by assistant versus
there internal meaning, was studied by Wiedijk \cite{zbMATH06319597}.)

We took care of this matter by providing, as an entry point for our
whole development, the theory \theory{Definitions\_Main} in which a
path from some the fundamental concepts from Isabelle/ZF reaching to our main theorems
is expounded. Cross-references to major milestones (which can be
navigated by using Isabelle) are provided there. It
corresponds roughly to less than seven journal pages.

Frequently, we formalized material by directly typing the proof we
new by heart, and in doing so we assumed that some definitions
accommodated some of our preconceptions.
It is significant that in a few such occasions, we were doubly
surprised by the fact that some supposedly trivial lemma would not go
through, because the definitions addressed something different (think
of \emph{restriction} of a function to a set versus that of a
relation), and also that we were able to proved the previous
results. The takeaway is that intuition may drive proofs
even if you are not working on what we believe we are.

A final aspect on this topic concerns automated methods. As it was
mentioned in the Introduction, some proofs can actually be
\emph{obscured} by automation, simply because one has to take
complicated opaque computations “on faith.”

@@

It is noteworthy that although the “math” of the construction of a
model of $\ZFC+\neg\CH$ was already in place by the end of November
2020, it was only 9 months later that we were able to finish the
formalization of that result. The missing pieces were essentially
bureacracy. Some of the material filed under this category comprises:
\begin{itemize}
\item calculation of arities of internalized formulas;
\item proving that certain constructions belong to the relevant
  models;
\item (required for the above) showing that particular instances of
  separation and replacement hold in the ground model;
\item etc.
\end{itemize}

Some of those proofs were almost copypasted once and again with minor
variants; this would usually be relegated to some function in the
meta-language, but we were unable to do this due to our limitations in
programming Isabelle/ML\footnote{%
  On the other hand, our inability to automate proofs of replacement
  instances paved the way for identifying which were the ones needed for
  forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
    place}).}
and  (an ML style having several quirks and which
is not very well document).

Nevertheless, the experience of the CS members of the team (and
community) is invaluable for the kind of engineering that is needed in
a large project like the present one. For example, it is
(mathematically) misleading when automatic tools (\isatt{simp},
\isatt{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeed but with 8
variables doesn't). Scale issues are very easily diregarded in the
abstract; for example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.

Another example concerns the very definition of $\forceisa$. As a
proof of concept, one of us tried to obtain its definition by
exclusively using formula synthesis, which was supposed to be trivial,
since it is mathematically. But in fact, some early minor mistake
rendered the whole effort useless. We then turned to a more
informed programming discipline, which involved decomposing the
definition in stages, each of which was checked for correctness, and
in that way we were able to reach our objective.
